Binning Strategies and Related Loss for Large Data

### 2015-03-09 Paper.Rnw
proofing

In the case of modestly sized data, scatterplots are great tools for showing relationships in two dimensions. For large data, scatterplots suffer from over-plotting -- i.e. \ktm{a large number of observations drawn in such close proximity that their points overlap}, thereby masking relevant structure.
Figure~\ref{fig:scatter-alpha} shows an example taken from baseball statistics. The scatterplot shows 139 seasons (from the years of 1871 -- 2009) of  pitching statistics for every baseball pitcher as published in Sean Lahman's Baseball database (\url{http://www.seanlahman.com/baseball-archive/}).  The number of games played in a season is plotted versus number of strikeouts a pitcher threw over the course of a season. While the data set is only medium sized with \Sexpr{dim(d1.ext)[1]} observations, it already shows some of the break-down patterns scatterplots experience with large data.

\ktm{ Figure~\ref{fig:scatter-alpha}a shows a traditional scatterplot with each observation is drawn with a filled circle. A triangular structure is apparent with some outliers at a medium number of games and high number of strikeouts; however the density within the triangular mass of points is indistinguishable}. Tukey \citep{tukey} suggested the use of open circles (see Figure~\ref{fig:scatter-alpha}b) to mitigate the problem of over-plotting. Open circles make points visible that are close together. This technique is  not suitable to determine the magnitude, but gives more information than is available with filled points. 
%\hh{We need to decide on one way for referring to $\alpha$-blending in one way - we've got about four different ways of writing it right now. I'm ok with any of them, but we need to stick to it.}
A modern alternative to open circles is alpha blending (see Figure~\ref{fig:scatter-alpha}c). Alpha blending provides more frequency information, allowing a distinction of higher-resolution density.

@@

\ktm{ Other scatterplot adaptations have been introduced that avoid overplotting by manipulating the display of the points by distorting the locations or the scales. Generalized scatterplots \citep{Keim2010GenScatter} display all individual observations but use distortion of the point locations by having points repel one another to avoid overlapping points.  An extension of generalized scatterplots uses clustering and local principal components to allow ellipsoid oriented distortion to display local correlation structure in the data \citep{Janetzko2013Ellipse}. Variable-binned scatterplots \citep{Hao2010visual} break the display into a non-uniform rectangular grid and resize the rows of cells according to density of points. This variable binning fragments the continuity of the axes into many segments on different scales and also does not deal with points that are exactly overlapped due to the resolution of data collection. Generalized and variable-binned scatterplots make fine data structure more visible and allows color to be reserved for a third variable instead of frequency; however, the distortion of the point locations or the axes wrap the visual display of the association between the two primary variables.}

\ktm{ Another approach is to reduce the graphical complexity by plotting binned aggregations of the data, namely frequencies, as opposed to plotting every observation as an individual point.} This has the additional advantage of reducing the size of the stored data \ktm{necessary for the construction of the plot}, as only the bin centers and the frequency of observations which correspond to that bin must be stored. \ktm{ Wickham argues for a ``bin-summarize-smooth'' procedure to be applied to the visualization of big data and he notes that simple summary functions, such as counts, scale very well with the size of data \citep{Wickham2013Bin}. Liu, Jiang and Heer employ the computational benefits of binning for their interactive big data visualization program, \textit{imMens} \citep{Liu2013imMens}. } \hh{also add \citep{gold} for binning reference - it's called `weighting' there.}

@@

%Histograms are an example of plots which use binned variables. \hh{XXX No, histograms don't use binned variables in their standard definition, histograms bin variables to render the information. Weighted histograms (again, \citet{gold}) make use of binned data, because we can think of the tabular information as the sufficient statistics for histograms.} 
\ktm{Histograms are a simple example of a plot that can be built using binned aggregations of the data; in their case the bin locations and bin counts act as a set of sufficient statistics necessary to reconstruct the plot. } A natural extensions of histograms to higher dimensions is to form a tessellated grid on a two dimensional Cartesian plane and use some other attribute, \hh{such as color or size (or 3D renderings, which we do not want to recommend at this point)} to provide joint density information within each grid cell, known as a tile. \ktm{ A {\it binned scatterplot} uses shading to provide frequency information, with tiles (rather than bars in a histogram) centered at the bin center, similar to a two-dimensional histogram viewed from above. A \textit{bubble plot} is a binned data plot that scales the size of a filled circle in proportion to frequency.  Bubble plots were first used by William Playfair \citep{playfair, playfair2}. }

@@

\ktm{  In general, in the univariate case with observations, $x_i$ for $i \in \{1,\dots,n\}$, binning algorithms require a set of bin centers $x_j^\ast$ for $j \in \{1,\dots,J\}$ and a function that maps observations to a bin centers. }

\ktm{ Rectangular binning accomplishes this by defining a sequence of $J$ adjacent intervals, $(\beta_{j-1},\beta_{j}]$ for $j \in \{1,\dots,J\}$, which span over the range of the data. Note that half open intervals are used such that any observation falling on a bin boundary is assigned to a unique interval. Values that fall at exactly $\beta_0$ are grouped into the first bin to close the leftmost bound. Each observation is then mapped to a bin center, $x_j^\ast$; the midpoint for the interval to which the observation belongs. }

This is expressed mathematically using the binning function $b_X(.) : x_i \rightarrow x^\ast_j$ is defined as 
%
\begin{eqnarray}\label{rectbin}
b_X(x_i) = \left\{\begin{array}{ll} 
  x^\ast_{1} &\text{ for all } x_i = \beta_{0} \\
  x^\ast_j & \text{ for all } x_i \in (\beta_{j-1} , \beta_j] 
  \end{array}\right.
\end{eqnarray}  
%
\textit{Standard rectangular binning} is a special cases of this general rectangular binning that uses intervals of equal size for all bins; thus only the origin of the first bin, $\beta_0$, and a binwidth, $\omega_X$, need to be specified. Standard rectangular binning is necessarily used in the construction of all histograms and is ideal for visualizing density because the consistency of binwidth makes the display of frequency proportional to density. Fixed width binning procedures are also highly computationally efficient \citep{Wickham2013Bin}.

Note that this binning procedure utilizes intervals that are open to the left and closes the outer bound of the leftmost bin. These specifications are consistent with the binning procedure used in the \texttt{hist()} function for creating histograms in base \texttt{R}. These specifications were selected for this paper, but these choices are by no means consider universal for binning.  For example, the \texttt{ggplot2} packages creates histograms with intervals open to the right and does not close the outer bound of the rightmost bin.

@@

As shown before, the overall loss partitions naturally into visual and numerical loss (see equation \ref{eqn:partition}). We can make use of this and
 recover some of the numerical information lost due to binning by storing the numerical center $(\bar{x},\bar{y})$ instead of the visual center $(x^\ast,y^\ast)$ for use in subsequent calculations. Storing the binned data as the arithmetic mean rather than the visual center reduces the total loss, and the visual center can be recovered from the bin width and  bin center $(O_X, O_Y)$, see table \ref{RecoveryComparison} for a numeric example for this.

>>>>

In the case of modestly sized data, scatterplots are great tools for showing relationships in two dimensions. For large data, scatterplots suffer from over-plotting, %-- i.e.\ a number of observations drawn in such close proximity that their points overlap, 
thereby masking relevant structure.
Figure~\ref{fig:scatter-alpha} shows an example taken from baseball statistics. The scatterplot shows 139 seasons (from the years of 1871 -- 2009) of  pitching statistics for every baseball pitcher as published in Sean Lahman's Baseball database (\url{http://www.seanlahman.com/baseball-archive/}).  The number of games played in a season is plotted versus number of strikeouts a pitcher threw over the course of a season. While the data set is only medium sized with \Sexpr{dim(d1.ext)[1]} observations, it already shows some of the break-down patterns scatterplots experience with large data.

Figure~\ref{fig:scatter-alpha}a shows a traditional scatterplot with each observation is drawn with a filled circle. A triangular structure is apparent with some outliers at a medium number of games and high number of strikeouts; however the density within the triangular mass of points is indistinguishable. \citet{tukey} suggested the use of open circles (see Figure~\ref{fig:scatter-alpha}b) to mitigate the problem of over-plotting. Open circles make points visible that are close together. This technique is  not suitable to determine the magnitude, but gives more information than is available with filled points. 
%\hh{We need to decide on one way for referring to $\alpha$-blending in one way - we've got about four different ways of writing it right now. I'm ok with any of them, but we need to stick to it.}
A modern alternative to open circles is alpha blending (see Figure~\ref{fig:scatter-alpha}c). Alpha blending provides more frequency information, allowing a distinction of higher-resolution density.

@@

Other scatterplot adaptations have been introduced that avoid overplotting by manipulating the display of the points by distorting the locations or the scales. Generalized scatterplots \citep{Keim2010GenScatter} display all individual observations (including ties) but use distortion of the point locations by having points repel one another to avoid overlapping points.  An extension of generalized scatterplots uses clustering and local principal components to allow ellipsoid oriented distortion to display local correlation structure in the data \citep{Janetzko2013Ellipse}. Variable-binned scatterplots \citep{Hao2010visual} break the display into a non-uniform rectangular grid and resize the rows of cells according to density of points. This variable binning fragments the continuity of the axes into  segments on different scales and also does not deal with ties. Generalized and variable-binned scatterplots make fine data structure more visible and allow color to be reserved for a third variable instead of frequency; however, the distortion of the point locations and/or axes warp the visual display of the association between the two primary variables.

Another approach is to reduce the graphical complexity by plotting binned aggregations of the data, namely frequencies, as opposed to plotting every observation as an individual point. This has the additional advantage of reducing the size of the stored data necessary for the construction of the plot, as only the bin centers and the frequency of observations which correspond to that bin must be stored. Wickham argues for a ``bin-summarize-smooth'' procedure to be applied to the visualization of big data and he notes that simple summary functions, such as counts, scale  well with the size of data \citep{Wickham2013Bin}. Liu, Jiang and Heer employ the computational benefits of binning for their interactive big data visualization program, \textit{imMens} \citep{Liu2013imMens}.  \hh{also add \citep{gold} for binning reference - it's called `weighting' there.}

@@

%Histograms are an example of plots which use binned variables. \hh{XXX No, histograms don't use binned variables in their standard definition, histograms bin variables to render the information. Weighted histograms (again, \citet{gold}) make use of binned data, because we can think of the tabular information as the sufficient statistics for histograms.} 
Histograms are a simple example of a plot that can be built using binned aggregations of the data; in their case the bin locations and bin counts act as a set of sufficient statistics necessary to reconstruct the plot.  A natural extensions of histograms to higher dimensions is to form a tessellated grid on a two dimensional Cartesian plane using some other attribute, \hh{such as color or size (or 3D renderings, which we do not want to recommend at this point)} to provide joint density information within each grid cell, known as a tile.  A {\it binned scatterplot} uses shading to provide frequency information, with tiles (rather than bars in a histogram) at the bin center, similar to a two-dimensional histogram viewed from above. A \textit{bubble plot} is a binned data plot that scales the size of a filled circle in proportion to frequency.  Bubble plots were first used by William Playfair \citep{playfair, playfair2}.

@@

In general, for the univariate case with observations, $x_i$ for $i \in \{1,\dots,n\}$, binning algorithms require a set of bin centers $x_j^\ast$ for $j \in \{1,\dots,J\}$ and a binning function $b_X(.) : x_i \rightarrow x^\ast_j$ that maps observations to  bin centers.

{\textit Rectangular binning} accomplishes this by defining a sequence of $J$ adjacent intervals, $(\beta_{j-1},\beta_{j}]$ for $j \in \{1,\dots,J\}$, which span over the range of the data. Note that half open intervals are used such that any observation falling on a bin boundary is assigned to a unique interval. Values $x_i$ exactly equal to the lowest bin boundar $\beta_0$ are grouped into the first bin to close the leftmost bound. Each observation is then mapped to a bin center, $x_j^\ast$; the midpoint for the interval to which the observation belongs.

% This is expressed mathematically using the binning function $b_X(.) : x_i \rightarrow x^\ast_j$ is defined as 
% %
% \begin{eqnarray}\label{rectbin}
% b_X(x_i) = \left\{\begin{array}{ll} 
%   x^\ast_{1} &\text{ for all } x_i = \beta_{0} \\
%   x^\ast_j & \text{ for all } x_i \in (\beta_{j-1} , \beta_j] 
%   \end{array}\right.
% \end{eqnarray}  
% %

\textit{Standard rectangular binning} is a special cases of general rectangular binning that uses intervals of equal size for all bins; thus only the origin of the first bin, $\beta_0$, and a binwidth, $\omega_X$, need to be specified. Standard rectangular binning is necessarily used in the construction of all histograms;  the consistent binwidth makes the display of frequency proportional to density. Fixed width binning procedures are also highly computationally efficient \citep{Wickham2013Bin}.

Note that this \hh{XXX which binning procedure are you referring to by 'this'?} binning procedure utilizes intervals that are open to the left and closes the outer bound of the leftmost bin. These specifications are consistent with the binning procedure used in the \texttt{hist()} function for creating histograms in base \texttt{R}. These specifications were selected for this paper, but these choices are by no means consider universal for binning.  For example, the \texttt{ggplot2} package creates histograms with intervals open to the right and does not close the outer bound of the rightmost bin.

@@

As shown before, the overall loss partitions naturally into visual and numerical loss (see equation \hh{XXX need new equation reference} %\ref{eqn:partition}
). We can make use of this and
 recover some of the numerical information lost due to binning by storing the numerical center $(\bar{x},\bar{y})$ instead of the visual center $(x^\ast,y^\ast)$ for use in subsequent calculations. Storing the binned data as the arithmetic mean rather than the visual center reduces the total loss, and the visual center can be recovered from the bin width and  bin center $(O_X, O_Y)$, see table \ref{RecoveryComparison} for a numeric example for this.
